---
title: "DATA 607: Project 3"
author: "Desiree Thomas, Denise Atherley, Kiera Griffiths"
format: html
editor: visual
---

## Approach -

This project is meant to explore the question, "**Which data science skills are most valued?**" in a collaborative process with a group of our choosing. The goal is to work effectively as a team to perform an exploratory analysis to answer this question. Because we've had experience working together in a past assignment and noticed that we collaborate pretty well, we (*Desiree Thomas, Denise Atherley and Kiera Griffiths*) decided to work together for this Project as well. We plan to fulfill the various project requirements by breaking down each key deliverable.

### Collaboration:

We will be using GitHub and Slack for our primary communication. GitHub will be the host for our shared repository and we will be using GitHub Projects to track issues, manage our workflow, assign tasks to specific members and track their completion. We will use Slack to coordinate meetings and Teams for video calls. GitHub issues may be used for handling bugs and errors. We have created a README.md for documentation as well.

### Data Acquisition:

To acquire a relevant data source, we used the assistance of LLM Google Gemini to suggest some data sources that could highlight which data science skills are most valued. Various options were suggested but we ultimately chose a data source from Kaggle titled, "[Data Science Job Postings with Salaries 2025](https://www.kaggle.com/datasets/elahehgolrokh/data-science-job-postings-with-salaries-2025)", which appeared to be the most compelling. The data is a processed version of scraped data collected in 2025 with transformations made to respect company privacy and avoid redistribution of raw proprietary content.

**Logical model for normalized database:**

To design a normalized relational schema, we will breakdown the Data Science Job Postings table into four interconnected tables.

-   **companies Table:** Stores unique information about each company.

    -   company_id (Primary Key)

    -   company_name (from the company column)

    -   headquarter, industry, ownership, company_size, revenue

-   **jobs Table:** Stores the specific job postings and links to the company.

    -   job_id (Primary Key)

    -   company_id (Foreign Key -\>companies.company_id)

    -   job_title, seniority_level, status, location, post_date, salary

-   **skills Table:** Stores a unique list of all possible skills.

    -   skill_id (Primary Key)

    -   skill_name

-   **job_skills Table:** A mapping (bridge) table to handle the many-to-many relationship between jobs and skills (since a job requires multiple skills, and a skill is required by multiple jobs).

    -   job_id (Foreign Key -\> jobs.job_id)

    -   skill_id (Foreign Key -\> skills.skill_id)

An Entity-Relationship (ER) diagram will be produced to document this design, illustrating how entities relate to one another. We can use resources such as draw.io, Excalidraw or Lucidchart.

**How we will load the data:**

We will load the .csv file into a database, most likely SQLite, and use SQL code to generate our structured tables with the appropriate primary keys so that we can link them.

Once the relational database is established, we can use the DBI, RSQLite and tidyverse libraries to perform all data tidying, transformations and exploratory data analysis in R. Hopefully, we can successfully store our database in the cloud which will allow our team to connect to it. We will pull only the data that we need for our analysis and address any data quality issues like missing data and outliers that deviate from the standard format.

### Analysis strategy:

In order to answer the question of which data science skills are most valued, we will likely analyze the data through two main lenses:

-   Market demand - what percentage of job postings require the specific skill?

-   Salary to skill correlation - which skills have the strongest positive correlation with the higher salary ranges.



```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(DBI)
library(RSQLite)
library(ggplot2)

# Load raw data and standardize headers
url <- "https://raw.githubusercontent.com/desithomas/607-Project-3/refs/heads/main/data_science_job_posts_2025.csv"

job_posts_raw <- read_csv(url) %>% 
  clean_names()
```



```{r}
job_posts_tidy <- job_posts_raw %>%
  # 1. Create the ID first
  mutate(job_id = row_number()) %>% 
  # 2. Clean the messy symbols
  mutate(salary_numeric = parse_number(salary)) %>%
  mutate(skills_clean = str_remove_all(skills, "[\\[\\]']")) %>%
  # 3. Convert text into calculable dates
  mutate(post_date_dt = today() - days(as.numeric(str_extract(post_date, "\\d+")))) %>%
  # 4. Filter out rows that are too messy to analyze
  drop_na(salary_numeric, skills_clean)


```


```{r}
jobs_table <- job_posts_tidy %>%
  select(job_id, job_title, salary_numeric, location)

skills_bridge <- jobs_tidy %>%
  select(job_id, skills_clean) %>%
  separate_rows(skills_clean, sep = ",\\s*")
```

```{r}
# Table A: Companies (Unique entities)
companies <- job_posts_tidy %>%
  select(company, headquarter, industry, company_size, revenue) %>%
  distinct(company, .keep_all = TRUE) %>% 
  mutate(company_id = row_number())

# Table B: Jobs Table (The Fact Table)
jobs_table <- job_posts_tidy %>%
  left_join(companies, by = "company") %>%
  select(job_id, job_title, salary_numeric, location, post_date_dt, company_id)

# Table C: Skills Bridge (Atomic data for analysis)
# This turns ['python', 'r'] into individual rows
skills_bridge <- job_posts_tidy %>%
  separate_rows(skills_clean, sep = ",\\s*") %>%
  mutate(skill_name = str_to_lower(str_trim(skills_clean))) %>%
  filter(skill_name != "") %>%
  mutate(skill_id = as.numeric(as.factor(skill_name))) %>%
  select(job_id, skill_id, skill_name)

# Table D: Skills Lookup (The Dictionary)
skills_lookup <- skills_bridge %>%
  select(skill_id, skill_name) %>%
  distinct()
```


```{r}
# 1. Create/Connect to the SQLite database file
con <- dbConnect(SQLite(), "ds_job_posts.db")

# 2. Write the four tables to the database
dbWriteTable(con, "companies", companies, overwrite = TRUE)
dbWriteTable(con, "jobs", jobs_table, overwrite = TRUE)
dbWriteTable(con, "skills_lookup", skills_lookup, overwrite = TRUE)
dbWriteTable(con, "job_skills_bridge", skills_bridge %>% select(job_id, skill_id), overwrite = TRUE)

# 3. Downstream Analysis: Querying "Value"
# We join the bridge to the jobs table to get the salary for each skill instance
final_analysis <- skills_bridge %>%
  left_join(jobs_table, by = "job_id") %>%
  group_by(skill_name) %>%
  summarise(
    demand = n(),
    avg_salary = round(mean(salary_numeric, na.rm = TRUE))
  ) %>%
  arrange(desc(avg_salary))

# Run a SQL query to check the cleaned salary
dbGetQuery(con, "SELECT job_title, salary_numeric FROM jobs LIMIT 10")

# Close connection
dbDisconnect(con)
```

```{r}
# Joining the Salary data to the Skills data
skill_valuation <- skills_bridge %>%
  # Bring in salary_numeric from the jobs_table
  left_join(jobs_table, by = "job_id") %>%
  
  # Group by each unique skill
  group_by(skill_name) %>%
  
  # Calculate Frequency (Demand) and Mean Salary (Value)
  summarise(
    demand_count = n(),
    avg_salary = round(mean(salary_numeric, na.rm = TRUE), 0)
  ) %>%
  
  # Sort by highest salary to see the most "Valued" skills
  arrange(desc(avg_salary))

# View the top 10 results
head(skill_valuation, 10)
```


```{r}

# Create the Value vs. Demand plot
ggplot(skill_valuation %>% filter(demand_count > 5), 
       aes(x = demand_count, y = avg_salary, label = skill_name)) +
  
  # Add the points
  geom_point(color = "darkgreen", size = 3, alpha = 0.7) +
  
  # Add text labels to the points (avoids overlapping)
  geom_text(vjust = -1, check_overlap = TRUE, size = 3) +
  
  # Format the Y-axis as currency
  scale_y_continuous(labels = scales::dollar_format()) +
  
  # Add professional titles
  labs(title = "Data Science Skills: Value vs. Demand (2025)",
       subtitle = "High-Value skills are at the top; High-Demand skills are to the right",
       x = "Demand (Number of Job Postings)",
       y = "Average Salary (Market Value)",
       caption = "Source: Project 3 Normalized SQLite Database") +
  
  theme_minimal()
```








